Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-26 agujournaltemplate.tex
Restored Ago's comments to go through them more carefully this time

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience and can therefore impact the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to unravel the effects on local communities. 
This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017. The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis. 
Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated with high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.

@@

    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term cyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods, data quality control, and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr. Kent Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{app:table1} in Appendix). 
    % Note: Perhaps add some more info about Supplmentary Results and test runs here?

@@

>>>>

\begin{abstract}
Phytoplankton communities play a crucial role in oceans' biogeochemical cycles. Climate drives changes in the physical environment that phytoplankton communities experience and can impact, thus, on the productivity and diversity of marine ecosystems. Long-term time series can capture changes in climate and ecosystems, allowing us to untangling of effects on local communities. 
This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017. The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data. 
Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated to high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004. However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009. To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors. Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.

@@

    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-termcyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods,  data quality control and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Kent Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

\subsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}. We extracted the -u wind component at 10\,m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). 
    % Note: Perhaps add some more info about Supplmentary Results and test runs here?

@@

The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
